---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This assignment uses two months of data from a personal activity monitoring device 
collected at 5 minute intervals throughout the day during October & November 2012.

## Loading and preprocessing the data

In step 1, the data file is loaded and prepared for analysis.

```{r chunk1, echo=TRUE, message=FALSE}
# unzip file
unzip(zipfile="./activity.zip", exdir="./")

# read in data
activity <- read.csv("./activity.csv")

# convert steps to numeric class
activity$steps <- as.numeric(activity$steps)

# load required package libraries
library(dplyr)

# place data into tbl_df container for the dplyr package
activitydata <- tbl_df(activity)
```
  
    

## What is mean total number of steps taken per day?

For step 2, the data is grouped by day to answer questions about total number of 
steps taken per day.

```{r chunk2, echo=TRUE}

# group data by date
DayGroup <- group_by(activitydata, date)

# calculate total number of steps per day
StepsPerDay <- summarise(DayGroup, steps = sum(steps, na.rm=TRUE))

# create histogram to show frequency of total number of steps taken per day
hist(StepsPerDay$steps, xlab="Steps Taken Per Day",
     main="Total Number of Steps Taken Per Day")

# set options for display of mean and median calculations below
options(scipen = 1, digits = 2)

# calculate the mean and median steps per day
MeanStepsPerDay <- mean(StepsPerDay$steps)
MedianStepsPerDay <- median(StepsPerDay$steps)
```

The mean steps taken each day is `r MeanStepsPerDay`  
The median steps taken each day is `r MedianStepsPerDay`  
  
    

## What is the average daily activity pattern?

Step 3 looks at the data grouped by 5 minute interval to find out the part of the
day when the most steps are taken.

```{r chunk3, echo=TRUE}

# group data by interval
IntervalGroup <- group_by(activitydata, interval)

# calculate average number of steps per interval
IntervalMeanSteps <- summarise(IntervalGroup, MeanSteps = mean(steps, na.rm=TRUE))

# create a plot showing average number of steps taken per interval
plot(IntervalMeanSteps, type="l", xlab="Interval", ylab="Average Number of Steps",
     main="Average Daily Activity Pattern")

# find the interval with the highest number of steps taken on average
max <- max(IntervalMeanSteps$MeanSteps)
maxInterval <- filter(IntervalMeanSteps, IntervalMeanSteps$MeanSteps==max)
```

5-minute interval with maximum number of steps on average: `r maxInterval$interval`  

## Imputing missing values

In step 4, missing values in the step data are imputed with the average number
of steps taken per interval, as calulated in step 3.

```{r chunk4, echo=TRUE}

# 4.1 - Calculate and report the total number of missing values in the dataset
TotalNAs <- sum(is.na(activitydata$steps))

# 4.2 - Devise a strategy for filling in all of the missing values in the dataset.

# My strategy is to impute values based on the average steps taken during each
# 5-min interval, as calculated in the last section.

# 4.3 - Create a new dataset that is equal to the original dataset but with
# the missing data filled in.

# create function to impute mean for NAs
NAforMean <- function(x) ifelse(is.na(x), mean(x, na.rm=TRUE), x)

# use mutate_each to impute by interval
activitydata2 <- activitydata %>%
        group_by(interval) %>%
        mutate_each(funs(NAforMean), steps)

# 4.4 - Make a histogram of the total number of steps taken each day and...
# calculate and report the mean and median total number of steps taken per day.

# group data from the new data set by date
DayGroup2 <- group_by(activitydata2, date)

# calculate the total number of steps taken each day 
StepsPerDay2 <- summarise(DayGroup2, steps = sum(steps))

# create a histogram showing total number of steps taken per day
hist(StepsPerDay2$steps, xlab="Steps Taken Per Day",
     main="Total Number of Steps Taken Per Day (w/Missing Data Imputed)")

# calculate the mean and median steps per day
MeanStepsPerDay2 <- mean(StepsPerDay2$steps)
MedianStepsPerDay2 <- median(StepsPerDay2$steps)
```

The mean steps taken each day is `r MeanStepsPerDay2`  
The median steps taken each day is `r MedianStepsPerDay2`  

## Are there differences in activity patterns between weekdays and weekends?

```{r chunk5, echo=TRUE}
# 5.1 - Create a new factor variable in the dataset with two levels – “weekday” and “weekend” 

# convert date to Date format
activitydata2$date2 <- as.Date(activitydata2$date, "%Y-%m-%d")

# return day of week from new Date
activitydata2$Weekday <- weekdays(activitydata2$date2)

# create a new factor variable called DayType 
activitydata2$DayType <- "weekday"
activitydata2$DayType[activitydata2$Weekday== "Saturday"] <- "weekend" 
activitydata2$DayType[activitydata2$Weekday== "Sunday"] <- "weekend" 

# 5.2 - Make a panel plot containing a time series plot

# group data by interval and DayType
IntervalGroup2 <- group_by(activitydata2, interval, DayType)

# calculate mean number of steps by groupings
IntervalMeanSteps2 <- summarise(IntervalGroup2, MeanSteps = mean(steps))

# set paramaters for panel plots
par(mfrow=c(2,1), mar=c(4,4,2,1)) 

# split data for plotting
weekdayData <- filter(IntervalMeanSteps2, DayType=="weekday")
weekdayData <- select(weekdayData, interval, MeanSteps)
weekendData <- filter(IntervalMeanSteps2, DayType=="weekend")
weekendData <- select(weekendData, interval, MeanSteps)

# create plots comparing average number of steps on weekdays vs weekends
plot(weekdayData, type="l", xlab="Interval", ylab="Number of Steps",
     main="weekday")
plot(weekendData, type="l", xlab="Interval", ylab="Number of Steps",
     main="weekend")
```




